---
title: "Random_Forest_Lab"
author: "Adriel Kim"
date: "12/1/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(rio)
library(plyr)
library(tidyverse)
library(rpart)
library(psych)
library(pROC)
#install.packages("rpart.plot")
library(rpart.plot)
#install.packages("rattle")
library(rattle)
library(caret)
library(C50) #Need this to pass into caret 
library(mlbench)
library(MLmetrics)

library(RColorBrewer)
library(ROCR)

library(mltools)
library(data.table)
library(randomForest)
library(help = randomForest)
library(rio)
```

The goal of this lab is to optimize a Random Forest model using the same dataset from last week and then compare the results from the C5.0 method to the Random Forest method. The guidance this week is less prescriptive in terms of steps, so use the skills you have gained over the semester to build and evaluate the RF model and determine if the RF is a better approach or not. You will be graded on your model building, interpretation of the results and explanation of model selection. As always, rely on your teams but submit your own code. Lastly, there are likely several correct approaches involving a variety of different conclusions, just make sure your conclusions are supported by your approach.    

The dataset below includes Census data on 32,000+ individuals with a variety of variables and a target variable for above or below 50k in salary. 

Your goal is to build a Random Forest Classifier to be able to predict income levels above or below 50k. 

```{r}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data"

census <- read_csv(url, col_names = FALSE)

colnames(census) <- c("age","workclass","fnlwgt","education","education_num","marital_status","occupation","relationship","race","sex","capital_gain","capital_loss","hours_per_week","native_country","income")


census$income <- as.factor(census$income)
View(census)

```


Recode the target variable to set the above 50k to 1 and below to 0, should already
be done. 
```{r}

levels(census$income) <- c(0, 1)
```

Ensure that the variables are correctly classified (should already be done)
```{r}

```

Finish any other data prep (one-hot encode, reduce factor levels)
```{r}
#one hot-encoding
#drop unneeded columns
#reduce factor levels by grouping other stuff
factors <- c("workclass", "marital_status", "occupation","relationship","race", "sex","native_country");
census[,factors] <- lapply(census[,factors], as.factor)
census_temp <- subset(census, select = -income)
census_hot <- one_hot(as.data.table(census_temp),cols = "auto",sparsifyNAs = TRUE,naCols = TRUE,dropCols = TRUE,dropUnusedLevels = TRUE) 

census_hot$income = census$income
names(census_hot) <- make.names(names(census_hot))

```

Create test, tune and training sets 
```{r}
#6 Split your data into test, tune, and train. (70/15/15)
part_index_1 <- caret::createDataPartition(census_hot$income,
                                           times=1,
                                           p = 0.70,
                                           groups=1,
                                           list=FALSE)

train <- census_hot[part_index_1, ]
tune_and_test <- census_hot[-part_index_1, ]

#The we need to use the function again to create the tuning set 

tune_and_test_index <- createDataPartition(tune_and_test$income,
                                           p = .5,
                                           list = FALSE,
                                           times = 1)

tune <- tune_and_test[tune_and_test_index, ]#aka, validation set
test <- tune_and_test[-tune_and_test_index, ]#final testing set


dim(train)
dim(test)# these will be slightly off because the data set isn't perfectly even
#buts its not a issue. 
dim(tune)
#View(train)
```

Calculate the initial mtry level 
```{r}
mytry_tune <- function(x){
  xx <- dim(x)[2]-1
  sqrt(xx)
}
mytry_tune(census_hot)
```

Run the initial RF model with 500 trees 
```{r}
census_RF = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 500,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 9,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 100,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF
```


Using the training and tune datasets tune the model in consideration of the number
of trees, the number of variables to sample and the sample size that optimize the model
output. 
```{r}
tune_test <- function(model){
  census_predict_tune = predict(model, tune, type="response", predict.all=FALSE, proximity = FALSE)
  (census_eval <- confusionMatrix(as.factor(census_predict_tune), 
                as.factor(tune$income), 
                dnn=c("Prediction", "Actual"), 
                mode = "sens_spec"))
}

census_RF_more_trees = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 1000,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 9,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 100,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF_more_trees

census_RF_greater_mtry = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 500,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 15,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 100,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF_greater_mtry

census_RF_more_samps = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 500,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 9,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 300,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF_more_samps

census_RF_more = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 1000,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 30,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 300,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF_more

tune_test(census_RF_more)

census_RF_more_tuning = randomForest(income ~ .,          #<- Formula: response variable ~ predictors.
                            #   The period means 'use all other variables in the data'.
                            train,     #<- A data frame with the variables to be used.
                            #y = NULL,           #<- A response vector. This is unnecessary because we're specifying a response formula.
                            #subset = NULL,      #<- This is unnecessary because we're using all the rows in the training data set.
                            #xtest = NULL,       #<- This is already defined in the formula by the ".".
                            #ytest = NULL,       #<- This is already defined in the formula by "PREGNANT".
                            ntree = 1000,        #<- Number of trees to grow. This should not be set to too small a number, to ensure that every input row gets classified at least a few times.
                            mtry = 50,            #<- Number of variables randomly sampled as candidates at each split. Default number for classification is sqrt(# of variables). Default number for regression is (# of variables / 3).
                            replace = TRUE,      #<- Should sampled data points be replaced.
                            #classwt = NULL,     #<- Priors of the classes. Use this if you want to specify what proportion of the data SHOULD be in each class. This is relevant if your sample data is not completely representative of the actual population 
                            #strata = NULL,      #<- Not necessary for our purpose here.
                            sampsize = 500,      #<- Size of sample to draw each time.
                            nodesize = 5,        #<- Minimum numbers of data points in terminal nodes.
                            #maxnodes = NULL,    #<- Limits the number of maximum splits. 
                            importance = TRUE,   #<- Should importance of predictors be assessed?
                            #localImp = FALSE,   #<- Should casewise importance measure be computed? (Setting this to TRUE will override importance.)
                            proximity = FALSE,    #<- Should a proximity measure between rows be calculated?
                            norm.votes = TRUE,   #<- If TRUE (default), the final result of votes are expressed as fractions. If FALSE, raw vote counts are returned (useful for combining results from different runs).
                            do.trace = TRUE,     #<- If set to TRUE, give a more verbose output as randomForest is run.
                            keep.forest = TRUE,  #<- If set to FALSE, the forest will not be retained in the output object. If xtest is given, defaults to FALSE.
                            keep.inbag = TRUE)   #<- Should an n by ntree matrix be returned that keeps track of which samples are in-bag in which trees? 

census_RF_more_tuning
tune_test(census_RF_more_tuning)

```

Once a final model has been selected, evaluate the model using the test dataset
```{r}

#Best results achieved with "census_RF_more_tuning" model with a validation accuracy of 0.8624
census_predict_test = predict(census_RF_more,      #<- a randomForest model
                            test,      #<- the test data set to use
                            type ="response",   #<- what results to produce, see the help menu for the options
                            predict.all = FALSE,  #<- should the predictions of all trees be kept?
                            proximity = FALSE)    #<- should proximity measures be computed

(census_eval <- confusionMatrix(as.factor(census_predict_test), 
                as.factor(test$income), 
                dnn=c("Prediction", "Actual"), 
                mode = "sens_spec"))


```


Summarize your findings as compared to the C5.0 model from last week. Think about the
time the model took to train, the model evaluation output and if the patterns generally 
between the two models are the same or different. What did you learn about the models or
the data along the way?

Compared to the decision tree model from last week, the random forest model had a better performance. However, it took longer to train especially with a large sample size and a large number of trees. Generally, the patterns of training and tuning the models were the same. Training and tuning both models involved creating multiple models and adjusting various hyperparameters. The tuning set allowed me to test my model's performance without peeking into the test data set. I would gradually adjust my hyperparameters based off validation accuracy. One thing I learned about the data was that the column names violated the naming format when one-hot-encoded. This required additional preprocessing in order for the data to work in the randomForest function.    
